\LaTeXe\ guide for the Journal of Functional Programming

### 2019-05-02 papers/jfp/8-engineering.tex
Rewrite S8.7

We can use the extra polymorphism from \S\ref{sec-polymorphism} for build tasks that produce multiple output
keys~--~for example, \cmd{ghc A.hs} produces both \cmd{A.hi} and \cmd{A.o}.
That can be represented by having a key whose value is a pair of file names, and
whose result is a pair of file contents. From that, the task for \cmd{A.hi}
can be the first component of the result of the pair. Such an operation
can be encoded without polymorphic keys provided the pair of files (or a dummy
file representing the pair) is marked as changed if either of the contained
files change, but it is more elegant and type-safe with the extra polymorphism.

Multiple-output build tasks require one more convenience mechanism,
which is orthogonal to polymorphism. Multiple-output task descriptions partition
the set of keys into \emph{elementary subsets} e.g.
$\{\cmd{A.hi},\cmd{A.o}\} \rightarrow \cmd{ghcTask}$,
$\{\cmd{B.o}\} \rightarrow \cmd{gccTask}$, etc. It is very convenient to
allow to fetch arbitrary sets of keys, such as $\{\cmd{A.o},\cmd{B.o}\}$,
without manually decomposing them into appropriate elementary subsets. This
requires a method to efficiently construct task descriptions operating on
arbitrary sets of keys rather on elementary subsets. One way to implement such
``multiple-output task closure'' is shown below.

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Partition k = k -> [k]

multi :: Eq k => Partition k -> Tasks Applicative [k] [v]
                             -> Tasks Applicative [k] [v]
multi partition tasks keys
    | k:_ <- keys, partition k == keys = tasks keys
    | otherwise = Just $ Task $ \fetch ->
        sequenceA [ select k <$> fetch (partition k) | k <- keys ]
  where
    select k = fromMaybe (error msg) . lookup k . zip (partition k)
    msg = "Partition invariants violated"
\end{minted}
\vspace{1mm}

\noindent
The function \hs{multi} takes a task description \hs{tasks} that operates on
elementary key subsets, and computes its closure supporting all possible
combinations of keys. The first argument \hs{partition} specifies the elementary
key subsets supported by the original \hs{tasks}. There are two cases: (i) if a
given set of \hs{keys} is an elementary subset, we pass the query to \hs{tasks};
(ii) otherwise, we examine every key \hs{k} in \hs{keys}, \hs{fetch} its
corresponding elementary subset, and \hs{select} the appropriate value from the
result. Returning to the above example, when given the query
$\{\cmd{A.o},\cmd{B.o}\}$, this function will \hs{fetch} elementary subsets
$\{\cmd{A.hi},\cmd{A.o}\}$ and $\{\cmd{B.o}\}$, thereby depending on the
corresponding build tasks \cmd{ghcTask} and \cmd{gccTask}.

We chose not to use polymorphic tasks in the above implementation for
simplicity, but nothing prevents us from combining polymorphism with
multiple-output task closure.

>>>>

Some build tasks produce multiple output keys -- for example \cmd{ghc A.hs} produces both
\cmd{A.hi} and \cmd{A.o}. This pattern can be encoded by having a key \cmd{A.hi+A.o} which produces
both results, then each of \cmd{A.hi} and \cmd{A.o} can extract the relevant result
from \cmd{A.hi+A.o}. We can express this pattern more clearly with the extra type safety from
\S\ref{sec-polymorphism}, using either pairs or lists for both the key and value:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
data Key a where
    File  :: FilePath -> Key String
    Files :: [FilePath] -> Key [String]

task :: TasksT Monad Key
task (File "A.hi") = Just $ TaskT $ \fetch -> do
    [hi, o] <- fetch (Files ["A.hi","A.o"])
    pure hi
task (File "A.o") = Just $ TaskT $ \fetch -> do
    [hi, o] <- fetch (Files ["A.hi","A.o"])
    pure o
task (Files ["A.hi","A.o"]) = Just $ TaskT $ \fetch ->
    -- run ghc here
\end{minted}
\vspace{1mm}

The only awkward aspect is that both \cmd{A.hi} and \cmd{A.o} must ask for exactly the same
\cmd{Files} key. If one \hs{File} key swapped the order of the list to \hs{Files} then it would likely
run \cmd{ghc} twice. To help construct a well-formed multiple-output build task
it is convenient to partition the set of keys into \emph{elementary subsets}.
We can encode such a partition with a function from any key to all the members of its
subset (in a consistent order).

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Partition k = k -> [k]
\end{minted}
\vspace{1mm}

\noindent In our example the partition function on either of the output names \hs{"A.hi"} or \hs{"A.o"} would
return \hs{["A.hi", "A.o"]}.

With a suitable \hs{Partition} it is possible to create a mapping that resolves \hs{File}
keys automatically into the correct \hs{Files} key.

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
task :: Partition FilePath -> TasksT Monad Key
task partition (File k) = Just $ TaskT $ \fetch -> do
    let ks = partition k
    let Just i = elemIndex k ks
    vs <- fetch (Files ks)
    pure (vs !! i)
-- patterns matching Files
\end{minted}
\vspace{1mm}

In the above function we compute the partition, find the index of this key within that
partition (using \hs{elemIndex}), run the \hs{fetch} on every key in the partition,
then extract out our specific result (using the indexing operation \hs{!!}).
